### 2024-01-05 sections/1-introduction.tex
general edits

Researchers are collecting data at an unprecedented scale.
In many fields, the sizes of datasets are growing exponentially, and this increase in the rate of data collection outpaces the rate of improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that computer architecture will not ``catch up'' to computational needs in the near future.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms for analyzing such datasets.

@@

For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions which do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could exhibit suboptimal classification accuracy in such contexts.

@@

This paper focuses on three novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison of CAKES to several state-of-the-art algorithms for similarity search, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also benchmark CAKES on a genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, the RadioML dataset~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.

@@

This search paradigm exploits the geometric and topological structure inherent in large datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with topological properties (such as the \textit{metric entropy} and \textit{local fractal dimension}, as defined in Section~\ref{sec:methods}) of the dataset, instead of its cardinality.
In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended entropy-scaling $\rho$-NN search from a flat clustering approach to a tree-based hierarchical clustering approach.
In this paper, we introduce CAKES (CLAM-Accelerated $K$-nn Entropy-scaling Search).
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the clustering algorithm from CHESS.
Using the cluster tree constructed by CLAM, CAKES extends CHESS to perform $k$-NN search and improves the performance of CHESS's $\rho$-NN search.

>>>>

Researchers are collecting data at an unprecedented scale.
In many fields, the sizes of datasets are growing exponentially, and this increase in the rate of data collection outpaces the rate of improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that computer architecture will not ``catch up'' to computational needs in the near future.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms for analyzing large datasets.

@@

For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions which do not obey the triangle inequality, such as Cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could exhibit suboptimal classification accuracy in such contexts.

@@

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-NN search algorithms in CHESS~\cite{ishaq2019clustered}, as well as improved genericity across distance functions.
We provide a comparison of CAKES's algorithms to several state-of-the-art algorithms for similarity search, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also benchmark CAKES on a genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, the RadioML dataset~\cite{oshea2018radioml} using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.

@@

This search paradigm exploits the geometric and topological structure inherent in large datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with topological properties (such as the \textit{metric entropy} and \textit{local fractal dimension}, as defined in Section~\ref{sec:methods}) of the dataset, instead of its cardinality.
In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended entropy-scaling $\rho$-NN search from a flat clustering approach to a tree-based hierarchical clustering approach.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the clustering algorithm from CHESS.
In this paper, we introduce CAKES, a set of three entropy-scaling algorithms for $k$-NN. 
Using the cluster tree constructed by CLAM, CAKES extends CHESS to perform $k$-NN search and improves the performance of CHESS's $\rho$-NN search.

### 2024-01-05 sections/2-methods.tex
fixed typo in methods and added to discussion
general edits

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among the points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Under distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics while Cosine distance is not a metric because it does not obey the triangle inequality (e.g. consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

@@

\begin{equation}
    \frac{\text{log} \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}

where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases.

We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, due to physical constraints (namely, biological evolution and the chemistry of RNA replication and mutation processes), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets obey the manifold hypothesis.

@@

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we would expect anything but a balanced tree.
The imbalance in the tree would be due to the varying density of the sampling in different regions of the manifold and the low dimensional ``shape'' of the manifold itself.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g. in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, we define the $k$-NN search problem which aims to find the $k$ closest points to $q$ in $\textbf{X}$;
i.e. the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$;
i.e. the set $H = B_X(q, \rho) = \{x \in \textbf{X}: f(q, x) \leq \rho \}$.

@@

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points in which are no more than a distance $\rho$ from the query.

@@

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \ 
        \underbrace{
            \overbrace{ \big| B_X(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}

where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

@@

where $\mu$ is the harmonic mean of the LFD of the clusters in $S$.
We use the harmonic mean to ensure that $\mu$ is not dominated by outlier clusters with very high LFD.
We cap the radial increase at 2 to ensure that we do not increase the radius too quickly in any single iteration.

>>>>

We define a \textit{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ which, given two points, deterministically returns a finite and non-negative real number.
A distance value of zero defines an identity among points (i.e. $f(x, y) = 0 \Leftrightarrow x = y$) and larger values indicate greater dissimilarity among the points.
We also require that the distance function be symmetric (i.e., $f(x, y) = f(y, x) \ \forall \ x, y \in \textbf{X}$).
In addition to these constraints, if the distance function obeys the triangle inequality (i.e. $f(x, y) \leq f(x, z) + f(z, y) \ \forall \ x, y, z \in \textbf{X}$), then it is also a \textit{distance metric}.
Similar to~\cite{yu2015entropy}, when used with distance metrics, all search algorithms in CAKES are exact.
For example, Euclidean, Levenshtein~\cite{levenshtein1966binary} and Dynamic Time Warping (DTW)~\cite{muller2007dynamic} distances are all distance metrics, while Cosine distance is not a metric because it violates the triangle inequality (e.g., consider the points $x = (1, 0)$, $y = (0, 1)$ and $z = (1, 1)$ on the Cartesian plane).

@@

\begin{equation}
    \frac{\text{log} \left( \frac{|B_X(q, r_1)|}{|B_X(q, r_2)|} \right) }{\text{log} \left( \frac{r_1}{r_2} \right) }
    \label{eq:methods:lfd-original}
\end{equation}
where $B_X(q, r)$ is the set of points contained in the metric ball of radius $r$ centered at a point $q$ in the dataset $\textbf{X}$.
Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases.

We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset which contains which contains genomes with unaligned lengths of up to 3,718 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,718 and at most 50,000.
However, due to physical constraints (namely, biological evolution and the chemistry of RNA replication and mutation processes), the data are constrained to a lower-dimensional manifold within this embedding space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Figure~\ref{fig:results:lfd-plots} illustrates this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.

@@

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we expect anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g. in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, we define the $k$-NN search problem, which aims to find the $k$ closest points to $q$ in $\textbf{X}$;
in other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.
We also have the $\rho$-NN search problem, which aims to find the all points in $\textbf{X}$ that are no more than a distance $\rho$ from $q$: the set $H = B_X(q, \rho) = \{x \in \textbf{X}: f(q, x) \leq \rho \}$.

@@

To perform $\rho$-NN search, we first perform a coarse \textit{tree-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:tree-search}, to find the leaf clusters which overlap with the query ball or any clusters which lie entirely within the query ball.
Then, for all such clusters, we perform a finer-grained \textit{leaf-search}, as outlined in Algorithm~\ref{alg:methods:rnn-search:leaf-search}, to find all points which are no more than a distance $\rho$ from the query.

@@

\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \ 
        \underbrace{
            \overbrace{ \big| B_X(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B_X(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

@@

where $\mu$ is the harmonic mean of the LFD of the clusters in $Q$.
We use the harmonic mean to ensure that $\mu$ is not dominated by outlier clusters with very high LFD.
We cap the radial increase at 2 to ensure that we do not increase the radius too quickly in any single iteration.

### 2024-01-05 sections/5-discussion.tex
general edits

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\mathcal{T} + \mathcal{L})$, where $\mathcal{T}$ is the time complexity of tree-search and $\mathcal{L}$ is the time complexity of leaf-search. 
Even though Repeated $\rho$-NN has the lowest time complexity (compare to $\mathcal{O}(\mathcal{T}\log{(\mathcal{T}+\mathcal{L})} + \mathcal{L}\log{(\mathcal{T}+\mathcal{L})})$ for Breadth-First Sieve and $\mathcal{O}(\mathcal{T}\log{\mathcal{T}} + \mathcal{L}\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically. 
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation. 
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD, 
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.  
This suggests that rather than use the reciprocal of the \textit{harmonic} mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may see better results by using the reciprocal of a mean which is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this overshooting occurs less frequently and with less severity.

>>>>

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{\mathcal{T}}{d} + \mathcal{L})$, where $\mathcal{T}$ is the time complexity of tree-search and $\mathcal{L}$ is the time complexity of leaf-search. 
Even though Repeated $\rho$-NN has the lowest time complexity (compare to $\mathcal{O}(\mathcal{T}\log{(\mathcal{T}+\mathcal{L})} + \mathcal{L}\log{(\mathcal{T}+\mathcal{L})})$ for Breadth-First Sieve and $\mathcal{O}(\mathcal{T}\log{\mathcal{T}} + \mathcal{L}\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically. 
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation. 
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD, 
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.  
This suggests that rather than use the reciprocal of the \textit{harmonic} mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may actually want a mean which is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this overshooting occurs less frequently and with less severity.
